Episodic memory: mental time travel or a quantum `memory wave' function?

### 2020-09-23 paper/main.tex
finished first pass of full revision addressing all R1 and R2 comments

One practical way to distinguish content from context representations is to separate mental (or neural) representations that drift rapidly \citep[content;][]{PolyEtal05a, MannEtal12} or gradually \citep[context;][]{PolyEtal05a, MannEtal11, HowaEtal12, LohnEtal18, LongKaha18, FolkEtal18}.  An intuition for why contextual representation might be expected to evolve gradually is laid out by \cite{PolyKaha08}.  Essentially, slowly drifting thoughts may be used to ``index'' more rapidly drifting thoughts that are temporally proximal.  This notion, that our brain maintains parallel mental representations drifting at different time scales, has been well-characterized in a series of elegant fMRI and ECoG studies by Uri Hasson and colleagues~\citep{HassEtal08, LernEtal11, HoneEtal12a, AlyEtal18}.  Subsequent work by the same group has shown that this hierarchy of drifting mental representations plays a central role in how we remember continuous experiences and segment our experiences into discrete events~\citep{BaldEtal17}.  This body of work suggests that different brain regions reflect \textit{temporal receptive windows} that characterize the timescale(s) to which that region is maximally sensitive.  This research complements work on the \textit{temporal window of integration} in the speech and perception literatures~\citep[e.g.,][]{ViemWake91, GautEtal12, vanWEtal07}.  The temporal window of integration refers to the duration over which physically distinct stimuli are treated (by some brain system or network) as a single percept or an atomic unit.  This systems-level work mirrors the discovery of \textit{time cells} in the rodent hippocampus that respond preferentially to a specific time relative to a temporal reference (e.g., time elapsed since starting to run on an exercise wheel), whereby each cell appears to integrate incoming information at a different rate~\citep{PastEtal08, MacDEtal11} and the population activity serves to represent information at a range of timescales~\citep{MauEtal18}. Other work suggests that the lateral entorhinal cortex also plays a role in integrating information across a wide range of timescales~\citep{TsaoEtal18}, supporting precise temporal recall~\citep{MontEtal19}.  These ideas have been formalized in a family of theoretical models over the past several decades~\citep{HowaKaha02a, DianEtal07, SedeEtal08, PolyEtal09, ShanEtal09, ShanHowa10, ShanHowa12, HowaEtal14, Rang18}.  In turn, these models were inspired in part by earlier theoretical work suggesting that temporally proximal experiences become linked through their shared contextual features~\citep{Este55a,AtkiShif68}.  Other recent work indicates that different aspects of the reinstatement process itself might also occur at multiple timescales.  For example, \cite{LindEtal19} found that high-level semantic information was reinstated prior to low-level (e.g. perceptual) details in an associative memory task.  Collectively, these experimental and theoretical studies make two important contributions to our understanding of how we remember our past.  First, the studies suggest that our ongoing experiences are ``blurred out'' in time by neural integrators (where different brain structures or sub-structures integrate with different time constants).  Second, the studies suggest that representations that reflect different timescales become bound together such that when we retrieve memories about our past, we reactivate representations that carry information at a range of time scales.  These reactivations explain how our brains reactivate prior contexts when we remember the past.

@@

\section*{Defining the ``quantum memory wave function'' metaphor and comparing it to the ``mental time travel'' metaphor}
Unlike in classical physics, where matter dutifully obeys Newton's laws, quantum physics postulates a different (though potentially complementary) set of rules governing how matter behaves and interacts.  Perhaps the most famous equation in quantum physics is \cite{Schr26}'s equation describing the quantum \textit{wave function}~\citep[this finding was also inspired by][]{deBr24}.  Schr\"{o}dinger's equation describes, in probabilistic terms, the location of a particle in space, at a given time.  In other words, it defines a function, $\Psi(x, t)$, such that the probability that a particle is located at position $x$ at time $t$ is equal to $|\Psi(x, t)|^2$.  The central idea is that, although it may not be possible to know or predict the particle's precise position at a given moment, the wave function tells us how it's location's probabilty mass extends over space.  In turn, this may allow one to use the wave function to predict behaviors or constraints of the system(s) in which the particle is behaving and interacting.  Within the domain of episodic memory, for the purposes of describing the \textit{quantum memory wave function} metaphor, we can conceptualize the moments along our \textit{autobiographical timeline} as the spatial coordinates ($x$) in Schr\"{o}dinger's equation.  Then our mental state at time $t$ may be conceptualized as spreading its probability mass over our autobiographical timeline.  In other words, like Schr\"{o}dinger's equation, we could in principle define a function describing how our mental state at time $t$ reflects the blends of thoughts we experienced at other moments of our lives.  Under this metaphor, the effect of mental operations we perform with respect to our autobiographical timeline, such as remembering past experiences or predicting what will happen in the future, is to shift the probability mass of our quantum memory wave function (Fig.~\ref{fig:qwave}).  This metaphor is also related to work on quantum probability decision models~\citep[e.g.,][]{KhreEtal18} in that it extends similar concepts to the domain of episodic memory.

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/wave_functions}
  \caption{\textbf{Memory wave functions. All panels.} Each panel displays a quantum memory wave function for a hypothetical individual at a particular moment (time $t=0$ indicated by $\blacktriangle$)).  The $x$-axes correspond to the position ($t$) along an autobiographical timeline.  The $y$-axes display how much the individual's thoughts at time $t = 0$ include information about the individual's past (or future) perceived or predicted experiences at other times. All curves have been normalized to have a maximum value of 1 and a minimum value of 0.  \textbf{A. Contextual drift (no explicit reinstatement).}  The blend of thoughts leading up to $t = 0$ reflects a recency-weighted average of information from the past.  \textbf{B. Mental time travel.} This constitutes a special case of the quantum memory wave function framework, whereby the individual's thoughts reflect a recency-weighted blend of prior thoughts, along with thoughts specific to one remembered prior experience (occuring at the timepoint indicated by the $\bigstar$ symbol).  Each curve's coloring denotes the amount of temporal blurring.  In this example, greater amounts of blurring denotes that the individual is bringing to mind information that is farther from the remembered moment or from $t = 0$.  \textbf{C. Quantum memory waves.}  In this example, the individual remembering the moment indicated by the $\bigstar$ symbol brings back a blend of thoughts similar to those shown in Panel B.  In addition, they bring to mind information pertaining to other conceptually related (but temporally non-contiguous) moments, indicated by $\star$ symbols. As in Panel B, different curves are colored according to their degree of temporal blurring.  \textbf{D. Generating predictions about the future.} To predict what will happen during a series of future events ($\diamond$ symbols) the individual spreads their thoughts over their prior experiences according to the expected relevance of those experiences to the future events.  Local peaks in the curve denote prior experiences with high expected relevance.  The left part of the panel (prior to $\blacktriangle$) illustrates how the individual's thoughts are spread out over their past, and the right part of the panel illustrates how the individual's thoughts are spread out over their (predicted) future.}
\label{fig:qwave}
\end{figure}

The \textit{mental time travel} metaphor is defined by Tulving as ``the type of awareness experienced when one thinks back to a \textit{specific moment} in one's personal past and consciusly recollects some prior episode or state as it was previously experienced''~\citep[Fig.~\ref{fig:qwave}B][emphasis added]{WheeEtal97}.  Whereas mental time travel refers to mentally revisiting to one specific moment in addition to now (potentially with some temporal blurring), the quantum memory wave framework implies that episodic memory entails spreading our awareness over \textit{many} moments (Fig.~\ref{fig:qwave}C).  In this way, the key distinction between these framings is whether remembering entails simultaneously revisiting one single moment versus many moments.  The two frameworks converge when we engage in \textit{chronesthesia}~\citep{Tulv02b}-- i.e., when we are consciously aware of subjective time.   Analogous to how measuring the physical location of a particle collapses its quantum wave function at the moment of measurement, engaging in chronesthesia can collapse our quantum memory wave function such that our thoughts temporarily converge on the moment we are revisiting.  This wave function expands again when our mental state changes in the next moment.

@@

@@

A rich recall density function analogous to the one displayed in Figure~\ref{fig:reinstatement}B could, in principle, be characteristic of memory for \textit{any} past event or experience.  For example, the peaks of the function could provide insight into which events, from which moments along one's autobiographical timeline, are functionally associated in memory.  Nonetheless, a potential challenge to testing this hypothesis in the laboratory is that in many memory studies there is no particular reason for participants to functionally associate the stimuli.  For example, there is often no ``deeper narrative'' to make sense of in list-learning studies or in most standard trial-based memory experiments.  However, random word list learning experiments might happen to contain (for a given list) semantically related words.  Indeed, a number of studies indicate that participants spontaneously form associations between semantically related words on random word lists, even when those related words are separated in time~\citep[e.g.,][]{WixtRohr94, MannKaha12, MannEtal12}.  This suggests that our memory systems pick up on statistical regularities, even in ostensibly ``random'' stimulus sequences, and that we leverage those regularities when we recall our past~\citep{PolyEtal09}.  Other work suggests that our memory systems may also leverage these regularities to parse our ongoing experiences into discrete events~\citep{SchaEtal13, Shap19}.  Are these examples in the literature on random list learning studies, indicating that our memory systems leverage ``accidental'' structure in random sequences, a reflection of the same fundamental principle driving the sorts of memory reactivation function displayed in Figure~\ref{fig:reinstatement}B?

@@

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} attempted to compute these embeddings using word co-occurance statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K~\times~K$ \textit{singular values} matrix, and a $K~\times$ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

@@

\textit{Topic models}, such as \textit{Latent Dirichlet Allocation}~\citep[LDA;][]{BleiEtal03} represent an attempt to characterize the latent \textit{topics} or themes that documents comprise (Fig.~\ref{fig:embedding-models}B).  Whereas matrix factorization approaches represent semantic meanings using low-dimensional approximations of word co-occurances or free association frequencies, topic models leverage these properties of documents (or behaviors) to explain \textit{why} words co-occur.  LDA defines a topic as a distribution of weights over words in the vocabulary.  For example, if one takes the \textit{vocabulary} to be the top 10,000 most commonly used words in English (excluding stop words that are devoid of explicit semantic meaning such as ``and,'' ``or,'' ``the,'' ``but,'' etc.), then each topic would comprise a vector of 10,000 numbers (weights) between zero and one, inclusive, where the full set of weights summed to one.  As shown in the cartoon on the left of the Panel (where the four top-weighted words from each of five topics are displayed), a fruit-related topic (shown in red) might weight words like ``apple,'' ``banana,'' ``pear,'' and ``orange'' heavily, whereas words like ``cat,'' ``dog,'' ``dolphin,'' and ``lobster'' might receive relatively small weights.  Given a set of $K$ topic vectors, topic models compute the blend of topics most probably reflected in a given document-- i.e., its \textit{topic proportions}.  These topic proportion estimates are made by examining how different topics weight each of the words in the document, and by assuming that documents tend to reflect few topics (rather than many topics).  Given only a word count matrix as input (just as in LSA), topic models automatically discover a $K$ by number-of-words \textit{topics matrix}, which describes how each topic (row) weights each word in the vocabulary (column), and a number-of-documents by $K$ \textit{topic proportions matrix}, which describes each document's topic proportions.  The rows of the topic proportions matrix may be treated as low-dimensional embeddings of the set of documents.

A key advantage that topic models have over matrix factorization models is the ability to explain potentially ambiguous word usage.  For example, consider the use of the word ``bat'' in the following illustrative passage:
\begin{quote}
  \textit{``As the slavering mutant vampire bat swooped low to attack, its eyes ablaze, the explorer grunted and took a mighty swing of her trusty aluminum baseball bat.  The powerful swing connected with a echoing metallic \textit{clink}, stunning the creature before it scrambled about on the cave floor and flew away.  She didn't even bat an eye-- as the Dark Wizard Chiroptera continued to draw strength from the mystic crystals, such things had become increasingly common of late, nearly to the point of mundanity...''}
  \end{quote}
At different ponts in the passage, ``bat'' refers to an animal, a piece of sports equipment, and a verb.  Because each of these uses of the word appear in the same document, models driven solely by word co-occurance statistics will blend these different meanings such that the embedding of ``bat'' is somewhat similar to other animals \textit{and} other pieces of sports equipment \textit{and} other eye-related verbs.  By contrast, treating each document as reflecting a blend of topics allows these uses to be disambiguated.  If other documents that mention the word ``bat'' describe vampires, swooping, or other mammals, but \textit{not} sports equipment or eye-related verbs, then a topic model could explain this observation by inferring that multiple topics place substantial weight on the word ``bat.''  Another strength of topic models is their interpretability.  Because the models directly output each document's topic proportions along with each topic's weights on all words in the vocabulary, the general themes that each topic reflects are often intuitive and the per-document topic proportions may be directly related to those themes.

@@

Many of these differences between physical and mental navigation also apply to other mental domains, including how we \textit{conceptualize} space (e.g., when we trace out a potential route in our mind).  Just as we mentally spread ourselves across a distribution of times when we remember, we can perform similar mental operations when we conceptualize and retrieve other types of information.  Further, these distributions may be estimated using existing methods.  For example, consider how we use pattern classifiers~\citep{NormEtal06b} to connect neural activity patterns to specific semantic~\citep{PolyEtal05a, MitcEtal08a, MannEtal12}, spatial~\citep{MillEtal13}, affective~\citep{ChanEtal18}, and social~\citep{MeyeEtal18} content being mentally manipulated.  A prevailing approach is to label specific brain activity patterns as each reflecting \textit{one particular} concept or mental state.  Variability in neural activity over repeated experiences or rememberings can result in classifiers spreading their predictions over multiple possible mental states, including non-contiguous mental states.  Traditional approaches then ``clean up'' this distribution over possible mental states by taking its average (expected), maximum, or modal value.  However, following the ``quantum wave function'' logic described in this manuscript, these multi-peaked receptive fields may provide a mechanism for associating non-contiguous representations.  What are some implications of this view for non-memory domains?

>>>>

One practical way to distinguish content from context representations is to separate mental (or neural) representations that drift rapidly \citep[content;][]{PolyEtal05a, MannEtal12} or gradually \citep[context;][]{PolyEtal05a, MannEtal11, HowaEtal12, LohnEtal18, LongKaha18, FolkEtal18}.  An intuition for why contextual representation might be expected to evolve gradually is laid out by \cite{PolyKaha08}.  Essentially, slowly drifting thoughts may be used to ``index'' more rapidly drifting thoughts that are temporally proximal.  This notion, that our brain maintains parallel mental representations drifting at different time scales, has been well-characterized in a series of elegant fMRI and ECoG studies by Uri Hasson and colleagues~\citep{HassEtal08, LernEtal11, HoneEtal12a, AlyEtal18}.  Subsequent work by the same group has shown that this hierarchy of drifting mental representations plays a central role in how we remember continuous experiences and segment our experiences into discrete events~\citep{BaldEtal17}.  This body of work suggests that different brain regions reflect \textit{temporal receptive windows} that characterize the timescale(s) to which that region is maximally sensitive.  This research complements work on the \textit{temporal window of integration} in the speech and perception literatures~\citep[e.g.,][]{ViemWake91, GautEtal12, vanWEtal07}.  The temporal window of integration refers to the duration over which physically distinct stimuli are treated (by some brain system or network) as a single percept or an atomic unit.  This systems-level work mirrors the discovery of \textit{time cells} in the rodent hippocampus that respond preferentially to a specific time relative to a temporal reference (e.g., time elapsed since starting to run on an exercise wheel), whereby each cell appears to integrate incoming information at a different rate~\citep{PastEtal08, MacDEtal11} and the population activity serves to represent information at a range of timescales~\citep{MauEtal18}. Other work suggests that the lateral entorhinal cortex also plays a role in integrating information across a wide range of timescales~\citep{TsaoEtal18}, supporting precise temporal recall~\citep{MontEtal19}.  In addition to this work suggesting that individual system or neurons maintain representations at their preferred timescales, a number of studies also report neurons and small populations exhibiting \textit{temporal multiplexing}, whereby information at different timescales is reflected in different features of an individual neuron's (or population's) activity patterns, such as spike timing versus firing rate and local field potential oscillations at different frequencies~\citep{CaruEtal18, KnigEich13, WatrEtal13, PanzEtal10}.

The above ideas about parallel neural representations that reflect information at multiple timescales have been formalized in a family of theoretical models over the past several decades~\citep{HowaKaha02a, DianEtal07, SedeEtal08, PolyEtal09, ShanEtal09, ShanHowa10, ShanHowa12, HowaEtal14, Rang18}.  In turn, these models were inspired in part by earlier theoretical work suggesting that temporally proximal experiences become linked through their shared contextual features~\citep{Este55a,AtkiShif68}.  Other recent work indicates that different aspects of the reinstatement process itself might also occur at multiple timescales.  For example, \cite{LindEtal19} found that high-level semantic information was reinstated prior to low-level (e.g. perceptual) details in an associative memory task.  Collectively, these experimental and theoretical studies make two important contributions to our understanding of how we remember our past.  First, the studies suggest that our ongoing experiences are ``blurred out'' in time by neural integrators (where different brain structures or sub-structures integrate with different time constants).  Second, the studies suggest that representations that reflect different timescales become bound together such that when we retrieve memories about our past, we reactivate representations that carry information at a range of time scales.  These reactivations explain how our brains reactivate prior contexts when we remember the past.

@@

\section*{Defining the ``quantum memory wave function'' metaphor and comparing it to the ``mental time travel'' metaphor}
Unlike in classical physics, where matter dutifully obeys Newton's laws, quantum physics postulates a different (though potentially complementary) set of rules governing how matter behaves and interacts.  Perhaps the most famous equation in quantum physics is \cite{Schr26}'s equation describing the quantum \textit{wave function}~\citep[this finding was also inspired by][]{deBr24}.  Schr\"{o}dinger's equation describes, in probabilistic terms, the location of a particle in space, at a given time.  In other words, it defines a function, $\Psi(x, t)$, such that the probability that a particle is located at position $x$ at time $t$ is equal to $|\Psi(x, t)|^2$.  The central idea is that, although it may not be possible to know or predict the particle's precise position at a given moment, the wave function tells us how it's location's probability mass extends over space.  In turn, this may allow one to use the wave function to predict behaviors or constraints of the system(s) in which the particle is behaving and interacting.  Within the domain of episodic memory, for the purposes of describing the \textit{quantum memory wave function} metaphor, we can conceptualize the moments along our \textit{autobiographical timeline} as the spatial coordinates ($x$) in Schr\"{o}dinger's equation.  Then our mental state at time $t$ may be conceptualized as spreading its probability mass over our autobiographical timeline.  In other words, like Schr\"{o}dinger's equation, we could in principle define a function describing how our mental state at time $t$ reflects the blends of thoughts we experienced at other moments of our lives.  Under this metaphor, the effect of mental operations we perform with respect to our autobiographical timeline, such as remembering past experiences or predicting what will happen in the future, is to shift the probability mass of our quantum memory wave function (Fig.~\ref{fig:qwave}).  This metaphor is also related to work on quantum probability decision models~\citep[e.g.,][]{KhreEtal18} in that it extends similar concepts to the domain of episodic memory.

\begin{figure}[tp] \centering \includegraphics[width=\textwidth]{figs/wave_functions}
  \caption{\textbf{Memory wave functions. All panels.} Each panel displays a quantum memory wave function for a hypothetical individual at a particular moment (time $t=0$ indicated by $\blacktriangle$)).  The $x$-axes correspond to the position ($t$) along an autobiographical timeline.  The $y$-axes display how much the individual's thoughts at time $t = 0$ include information about the individual's past (or future) perceived or predicted experiences at other times. All curves have been normalized to have a maximum value of 1 and a minimum value of 0.  \textbf{A. Contextual drift (no explicit reinstatement).}  The blend of thoughts leading up to $t = 0$ reflects a recency-weighted average of information from the past.  \textbf{B. Mental time travel.} This constitutes a special case of the quantum memory wave function framework, whereby the individual's thoughts reflect a recency-weighted blend of prior thoughts, along with thoughts specific to one remembered prior experience (occurring at the timepoint indicated by the $\bigstar$ symbol).  Each curve's coloring denotes the amount of temporal blurring.  In this example, greater amounts of blurring denotes that the individual is bringing to mind information that is farther from the remembered moment or from $t = 0$.  \textbf{C. Quantum memory waves.}  In this example, the individual remembering the moment indicated by the $\bigstar$ symbol brings back a blend of thoughts similar to those shown in Panel B.  In addition, they bring to mind information pertaining to other conceptually related (but temporally non-contiguous) moments, indicated by $\star$ symbols. As in Panel B, different curves are colored according to their degree of temporal blurring.  \textbf{D. Generating predictions about the future.} To predict what will happen during a series of future events ($\diamond$ symbols) the individual spreads their thoughts over their prior experiences according to the expected relevance of those experiences to the future events.  Local peaks in the curve denote prior experiences with high expected relevance.  The left part of the panel (prior to $\blacktriangle$) illustrates how the individual's thoughts are spread out over their past, and the right part of the panel illustrates how the individual's thoughts are spread out over their (predicted) future.}
\label{fig:qwave}
\end{figure}

The \textit{mental time travel} metaphor is defined by Tulving as ``the type of awareness experienced when one thinks back to a \textit{specific moment} in one's personal past and consciously recollects some prior episode or state as it was previously experienced''~\citep[Fig.~\ref{fig:qwave}B][emphasis added]{WheeEtal97}.  Whereas mental time travel refers to mentally revisiting to one specific moment in addition to now (potentially with some temporal blurring), the quantum memory wave framework implies that episodic memory entails spreading our awareness over \textit{many} moments (Fig.~\ref{fig:qwave}C).  In this way, the key distinction between these framings is whether remembering entails simultaneously revisiting one single moment versus many moments.  The two frameworks converge when we engage in \textit{chronesthesia}~\citep{Tulv02b}-- i.e., when we are consciously aware of subjective time.   Analogous to how measuring the physical location of a particle collapses its quantum wave function at the moment of measurement, engaging in chronesthesia can collapse our quantum memory wave function such that our thoughts temporarily converge on the moment we are revisiting.  This wave function expands again when our mental state changes in the next moment.

@@

While the studies presented in the \textit{Introduction and overview} on temporal receptive windows, the temporal window of integration, time cells, and temporal multiplexing present strong empirical evidence that our brains maintain parallel representations at multiple timescales, questions remain about how or whether we experience those representations at a conscious level.  For example, are we consciously aware of spreading our mental state over many moments from our past?  Or is our subjective experience of remembering more like mentally time traveling back to a single ``hybrid'' moment that blends information from several of our (actual) prior experiences?

@@

A rich recall density function analogous to the one displayed in Figure~\ref{fig:reinstatement}B could, in principle, be characteristic of memory for \textit{any} past event or experience.  For example, the peaks of the function could provide insight into which events, from which moments along one's autobiographical timeline~\citep{ArzyEtal09}, are functionally associated in memory.  Nonetheless, a potential challenge to testing this hypothesis in the laboratory is that in many memory studies there is no particular reason for participants to functionally associate the stimuli.  For example, there is often no ``deeper narrative'' to make sense of in list-learning studies or in most standard trial-based memory experiments.  However, random word list learning experiments might happen to contain (for a given list) semantically related words.  Indeed, a number of studies indicate that participants spontaneously form associations between semantically related words on random word lists, even when those related words are separated in time~\citep[e.g.,][]{WixtRohr94, MannKaha12, MannEtal12}.  This suggests that our memory systems pick up on statistical regularities, even in ostensibly ``random'' stimulus sequences, and that we leverage those regularities when we recall our past~\citep{PolyEtal09}.  Other work suggests that our memory systems may also leverage these regularities to parse our ongoing experiences into discrete events~\citep{SchaEtal13, Shap19}.  Are these examples in the literature on random list learning studies, indicating that our memory systems leverage ``accidental'' structure in random sequences, a reflection of the same fundamental principle driving the sorts of memory reactivation function displayed in Figure~\ref{fig:reinstatement}B?

@@

Early word embedding models such as \textit{Latent Semantic Analysis}~\citep[LSA;][]{DeerEtal90, LandDuma97} attempted to compute these embeddings using word co-occurence statistics.  The logic driving these approaches posited that documents that contained similar mixtures of words would be likely to be conceptually similar, and words that appeared in similar documents would also likely be conceptually similar.  To quantify these intuitions, \cite{LandDuma97} first constructed \textit{word count matrices} (Fig.~\ref{fig:embedding-models}A) for several text corpora (each comprising many thousands of documents).  Each entry in these matrices denotes the number of time a given document (row) contained a given word (column).  They then used the Singular Value Decomposition (SVD) to decompose the number-of-documents by number-of-words word count matrix into the product of a number-of-documents $\times~K$ \textit{document embeddings matrix}, a $K~\times~K$ \textit{singular values} matrix, and a $K~\times$ number-of-unique-words \textit{word embeddings matrix}.  The rows of the document embeddings matrix reflected the similarities between documents, and the columns of the word embeddings matrix reflected similarities between words.

@@

\textit{Topic models}, such as \textit{Latent Dirichlet Allocation}~\citep[LDA;][]{BleiEtal03} represent an attempt to characterize the latent \textit{topics} or themes that documents comprise (Fig.~\ref{fig:embedding-models}B).  Whereas matrix factorization approaches represent semantic meanings using low-dimensional approximations of word co-occurrences or free association frequencies, topic models leverage these properties of documents (or behaviors) to explain \textit{why} words co-occur.  LDA defines a topic as a distribution of weights over words in the vocabulary.  For example, if one takes the \textit{vocabulary} to be the top 10,000 most commonly used words in English (excluding stop words that are devoid of explicit semantic meaning such as ``and,'' ``or,'' ``the,'' ``but,'' etc.), then each topic would comprise a vector of 10,000 numbers (weights) between zero and one, inclusive, where the full set of weights summed to one.  As shown in the cartoon on the left of the Panel (where the four top-weighted words from each of five topics are displayed), a fruit-related topic (shown in red) might weight words like ``apple,'' ``banana,'' ``pear,'' and ``orange'' heavily, whereas words like ``cat,'' ``dog,'' ``dolphin,'' and ``lobster'' might receive relatively small weights.  Given a set of $K$ topic vectors, topic models compute the blend of topics most probably reflected in a given document-- i.e., its \textit{topic proportions}.  These topic proportion estimates are made by examining how different topics weight each of the words in the document, and by assuming that documents tend to reflect few topics (rather than many topics).  Given only a word count matrix as input (just as in LSA), topic models automatically discover a $K$ by number-of-words \textit{topics matrix}, which describes how each topic (row) weights each word in the vocabulary (column), and a number-of-documents by $K$ \textit{topic proportions matrix}, which describes each document's topic proportions.  The rows of the topic proportions matrix may be treated as low-dimensional embeddings of the set of documents.

A key advantage that topic models have over matrix factorization models is the ability to explain potentially ambiguous word usage.  For example, consider the use of the word ``bat'' in the following illustrative passage:
\begin{quote}
  \textit{``As the slavering mutant vampire bat swooped low to attack, its eyes ablaze, the explorer grunted and took a mighty swing of her trusty aluminum baseball bat.  The powerful swing connected with a echoing metallic \textit{clink}, stunning the creature before it scrambled about on the cave floor and flew away.  She didn't even bat an eye-- as the Dark Wizard Chiroptera continued to draw strength from the mystic crystals, such things had become increasingly common of late, nearly to the point of mundanity...''}
  \end{quote}
At different ponts in the passage, ``bat'' refers to an animal, a piece of sports equipment, and a verb.  Because each of these uses of the word appear in the same document, models driven solely by word co-occurrence statistics will blend these different meanings such that the embedding of ``bat'' is somewhat similar to other animals \textit{and} other pieces of sports equipment \textit{and} other eye-related verbs.  By contrast, treating each document as reflecting a blend of topics allows these uses to be disambiguated.  If other documents that mention the word ``bat'' describe vampires, swooping, or other mammals, but \textit{not} sports equipment or eye-related verbs, then a topic model could explain this observation by inferring that multiple topics place substantial weight on the word ``bat.''  Another strength of topic models is their interpretability.  Because the models directly output each document's topic proportions along with each topic's weights on all words in the vocabulary, the general themes that each topic reflects are often intuitive and the per-document topic proportions may be directly related to those themes.

@@

Many of these differences between physical and mental navigation also apply to other mental domains, including how we \textit{conceptualize} space~\citep[e.g., when we trace out a potential route in our mind;][]{GautvanW16, GautEtal18, ArzyScha19}.  Just as we mentally spread ourselves across a distribution of times when we remember, we can perform similar mental operations when we conceptualize and retrieve other types of information.  Further, these distributions may be estimated using existing methods.  For example, consider how we use pattern classifiers~\citep{NormEtal06b} to connect neural activity patterns to specific semantic~\citep{PolyEtal05a, MitcEtal08a, MannEtal12}, spatial~\citep{MillEtal13}, affective~\citep{ChanEtal18}, and social~\citep{MeyeEtal18} content being mentally manipulated.  A prevailing approach is to label specific brain activity patterns as each reflecting \textit{one particular} concept or mental state.  Variability in neural activity over repeated experiences or rememberings can result in classifiers spreading their predictions over multiple possible mental states, including non-contiguous mental states.  Traditional approaches then ``clean up'' this distribution over possible mental states by taking its average (expected), maximum, or modal value.  However, following the ``quantum wave function'' logic described in this manuscript, these multi-peaked receptive fields may provide a mechanism for associating non-contiguous representations.  What are some implications of this view for non-memory domains?
